Computational methods in landscape ecology

### 2024-10-19 Sections/01-introduction.qmd
Clean changes MJ

Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, but are not limited to, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modeling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
<!-- R1: Data-driven and model-driven -->
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven and model-driven approaches [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of experimental studies].
Additionally, because landscape ecology is a cross-disciplinary field [@With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

<!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
Here, we introduce recent computational developments in multiple key topics of landscape ecology (Fig. 1), including software that implements or is potentially capable of executing them (Tab. 1).
In this context, we are focusing on open-source software and scripting languages such as R, Python, and Julia.
However, we do not provide a systematic literature review or a general introduction to (computational) landscape ecology.
For such, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.

>>>>

Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land-use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, but are not limited to, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modeling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
<!-- R1: Data-driven and model-driven -->
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven and model-driven approaches [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of experimental studies].
Additionally, because landscape ecology is a cross-disciplinary field [@With2019EssentialsLandscape], the availability of data increases steadily [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is a need for advanced computational methods.

<!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
Here, we provide a perspective on recent developments and advances in computational methods in multiple key topics of landscape ecology (Fig. 1), including software that implements them or is potentially capable of creating novel insights in these topics (Tab. 1).
In this context, we are focusing on open-source software and scripting languages such as R, Python, and Julia.
However, we do not aim to provide a systematic literature review or a general introduction to (computational) landscape ecology.
For such, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.

### 2024-10-19 Sections/02-data.qmd
Clean changes MJ

Similarly the issue of scale is closely related to the used data model. This includes the extent of the study area and resolution of the data (i.e., the smallest data unit).
Additionally, thematic resolution in landscape ecology often refers to the values describing landscape features, e.g., the number of discrete LULC categories or habitats [@Jung2020global].
Various classification systems are relevant at local, regional, and global extent, such as the National Land Cover Database (NLCD), the Coordination of Information on the Environment (CORINE), or the Food and Agriculture Organization (FAO) systems [@Nedd2021SynthesisLand].
Lastly, temporal scales are relevant in terms of extent and resolution, but also with regards to the detail of captured processes acting at the landscape scale.

The spatial reference system of the data is less frequently discussed but often critical. 
Cartesian coordinate reference systems measure distances between two points in Euclidean distances, typically in meters. 
In contrast, geographic coordinate reference systems are based on degrees and measure distances as great-circle distances.
Numerous coordinate reference systems exist and novel projections are constantly being developed to improve spatial cohesion, efficiency and accuracy [@bauer2014optimisation]. However each reference systems comes with distinct properties that may distort areas, distances, angles, and aimed to represent specific regions or countries. 
Therefore, for analysts it is essential to choose the appropriate spatial reference system based on the available data, area of interest, and the specific research question.

Data quality is generally variable, and data products are not flawless, potentially containing errors and biases.
For example, global land-cover products typically have an overall accuracy of 70 to 80% [@xu2024comparative], indicating that 20 to 30% of the grid cells may be misclassified.
Furthermore, these misclassifications are not random, but are often correlated with specific LULC types, regions, and seasonal variations [@Liu2021FinerresolutionMapping]. Rather than relying on global products, there are however also increasingly voices that advocate for the production and use of regional land-cover products as alternative [@tulbure2022regional].
Nevertheless, many landscape ecology studies tend to accept the data at face value, neglecting its accuracy and inherent uncertainty.

>>>>

Similarly the issue of scale is closely related to the used data model. 
This includes the extent of the study area and resolution of the data (i.e., the smallest data unit).
Additionally, thematic resolution in landscape ecology often refers to the values describing landscape features, e.g., the number of discrete LULC categories or habitats
[@Jung2020GlobalMap].
Various classification systems are relevant at local, regional, and global extent, such as the National Land Cover Database (NLCD), the Coordination of Information on the Environment (CORINE), or the Food and Agriculture Organization (FAO) systems [@Nedd2021SynthesisLand].
Lastly, temporal scales are relevant in terms of extent and resolution, but also with regards to the detail of captured ecological processes.

The spatial reference system of the data is less frequently discussed but often critical. 
Cartesian coordinate reference systems measure distances between two points in Euclidean distances, typically in meters. 
In contrast, geographic coordinate reference systems are based on degrees and measure distances as great-circle distances.
Numerous coordinate reference systems exist and novel projections are constantly being developed to improve spatial cohesion, efficiency and accuracy [@Bauer-Marschallinger2014OptimisationGlobal].
However each reference systems comes with distinct properties that may distort areas, distances, angles, and aimed to represent specific regions or countries. 
Therefore, it is essential to choose the appropriate spatial reference system based on the available data, area of interest, and the specific research question.

Data quality is generally variable, and data products are not flawless, potentially containing errors and biases.
For example, global land-cover products typically have an overall accuracy of 70 to 80% [@Xu2024ComparativeValidation], indicating that 20 to 30% of the grid cells may be misclassified.
Furthermore, these misclassifications are not random, but are often correlated with specific LULC types, regions, and seasonal variations [@Liu2021FinerresolutionMapping]. 
Rather than relying on global products, however, there are also increasingly voices that advocate for the production and use of regional land-cover products [@Tulbure2022RegionalMatters].
Nevertheless, many landscape ecology studies tend to accept the data at face value, neglecting its accuracy and inherent uncertainty.

### 2024-10-19 Sections/03-spatial-patterns.qmd
Clean changes MJ

Traditionally and at present, many prominent approaches that quantify spatial patterns revolve around raster data using categorical values based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
<!-- R1: They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time -->
The strength of landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data based on remote sensing derived products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
However, also limitations have been identified for many landscape metrics. 
These limitations include shortcomings to quantify the spatial structure, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [@Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

@@

The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape, omitting the spatial configuration.
Shannon's entropy can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and interclass distances [@Claramunt2012SpatiotemporalForm].
However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory must be adapted.
@Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix which could be used to calculate various entropy measures.
This includes conditional entropy (representing complexity in configuration), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).

Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, in the last few years, there has been a surge in the development of entropy-based metrics for landscape ecology.
In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of 'macrostate' (the general state of a system) and 'microstate' (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
<!-- R1: the edge length -> Could you specify a bit -->
@Cushman2016CalculatingConfigurational proposed to relate the edge length (defined as the side lengths of neighboring cells with different LULC classes) to the microstate of the landscape and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
The relationship between Shannon and Boltzmann entropies in landscape ecology remains contentious, with recent studies challenging the thermodynamic interpretation of Boltzmann-inspired measures and advocating for Shannon entropy as a more general form [@Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial], highlighting the need for further research to elucidate their connections to environmental processes.

>>>>

Traditionally and at present, many prominent approaches that quantify spatial patterns revolve around raster data using categorical values based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
<!-- R1: They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time -->
The strength of landscape metrics is that they are easy to apply, communicate, and can be calculated straightforwardly from raster data based on remote sensing derived products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
However, also limitations have been identified for many landscape metrics.
These limitations include shortcomings to quantify the spatial structure, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [@Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

@@

The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape, omitting the spatial configuration.
Shannon's entropy can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and interclass distances [@Claramunt2012SpatiotemporalForm].
However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory must be adapted.
@Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix which could be used to calculate various entropy measures.
This includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).

Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, in the last few years, there has been a surge in the development of entropy-based metrics for landscape ecology.
In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of "macrostate" (the general state of a system) and "microstate" (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
<!-- R1: the edge length -> Could you specify a bit -->
@Cushman2016CalculatingConfigurational proposed to relate the edge length (defined as the side lengths of neighboring cells with different LULC classes) to the microstate of the landscape and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
The relationship between Shannon and Boltzmann entropies in landscape ecology remains contentious, with recent studies challenging the thermodynamic interpretation of Boltzmann-inspired measures and advocating for Shannon entropy as a more general form [@Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial], highlighting the need for further research to elucidate their connections to environmental processes.

### 2024-10-19 Sections/05-sampling.qmd
Clean changes MJ

Furthermore, identifying scales at which the ecological processes of interest operate is a crucial step [@Jackson2015AreEcologists].
However, in practice, the appropriate scales may not be obvious or data at such scales may not be available and landscape-wide and local process interactions can occur across multiple scales [@miguet2016determines].
Here analyses at multiple spatial scales can help to identify the scales of interest. For instance, scalograms can be used to reveal scale thresholds that maximize landscape heterogeneity [@Pasher2013OptimizingLandscape] or show the strongest relationships with response variables [@Huais2024MultilandrPackage].

>>>>

Furthermore, identifying scales at which the ecological processes of interest operate is a crucial step [@Jackson2015AreEcologists].
However, in practice, the appropriate scales may not be obvious or data at such scales may not be available and landscape-wide and local process interactions can occur across multiple scales [@Miguet2016WhatDetermines].
Here analyses at multiple spatial scales can help to identify the scales of interest.
For instance, scalograms can be used to reveal scale thresholds that maximize landscape heterogeneity [@Pasher2013OptimizingLandscape] or show the strongest relationships with response variables [@Huais2024MultilandrPackage].

### 2024-10-19 Sections/06-simulations.qmd
Clean changes MJ

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling]. 
They can also incorporate a range of socio-economic, political and governance information to influence how agents (such as farmers) interact within a landscape [@murray2014combining].
In order to facilitate future model developments recent progress includes adaptable modeling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches, coupling or integrating different model types with individual-based models, could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

@@

The digital twins concept include three connected elements, namely *i)* a physical object or system, *ii)* its digital representation, *iii)* and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins]. Digital twins are particular useful for real-time workflows that allow now- and forecasting of complex dynamics in a landscape.
The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
They aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].
<!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g. drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
This is further facilitated by the increasing rate of novel sources of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].

Thus, there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
Digital twins are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins]. Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin] or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.

>>>>

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling].
They can also incorporate a range of socio-economic, political, and governance information to influence how individuals, such as farmers, interact within a landscape [@Murray-Rust2014CombiningAgent].
In order to facilitate future model developments recent progress includes adaptable modeling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches, coupling or integrating different model types with individual-based models, could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

@@

The digital twins concept include three connected elements, namely *i)* a physical object or system, *ii)* its digital representation, *iii)* and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins].
Digital twins are particular useful for real-time workflows that allow now- and forecasting of complex dynamics in a landscape.
The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology.
They aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].
Thus, there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
Digital twins are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins]. 
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin] or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
<!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g. drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
This is further facilitated by the increasing rate of novel sources of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].

### 2024-10-19 Sections/08-spatial-planning.qmd
Clean changes MJ

Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
In the context of landscape ecology, it provides a pattern-process link and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology.
For example, there have been several recent advances in considering landscape patterns directly in spatial planning or through using simulation model outputs as input features [@Daigle2020OperationalizingEcological].

@@

A second development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
<!-- R1: Beware, deep reinforcement learning is not an exact approach, and thus cannot provide guarantees that the best possible solution has been found. Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
However, this comes with the drawbacks of reduced interpretability and because it is not an exact approach no guarantee that the best possible solution can been found.

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
Most software are based on linear or constrained programming approaches and require a solver to create any outputs.
<!-- R1: Be careful with such statements, it is extremely complex to objectively compare two solvers, and almost impossible to state the general superiority of one state-of-the-art solver against another one. Indeed, some solvers may be very efficient for some kinds of problems while very inefficient for others. Also, modelling and solver parametrization can make a lot of differences in solving efficiency and has often a lot of impact on performances. They are solver competitions, however, they are based on benchmarks, therefore classic and well-known problems that do not always reflect common uses of solvers. A solver fine-tuned to be good on such benchmarks will not necessarily perform well on any problem. There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
Unfortunately, there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing] and the optimal solver will depend on the specific problem. 
However, due to software licenses and related cost the optimal solver might not be always accessible.

>>>>

Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState; @Hanson2024SystematicConservation].
In the context of landscape ecology, it provides a pattern-process link and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology.
For example, there have been several recent advances in considering landscape patterns directly in spatial planning or through using simulation model outputs as input features [@Daigle2020OperationalizingEcological].

@@

A second development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
<!-- R1: Beware, deep reinforcement learning is not an exact approach, and thus cannot provide guarantees that the best possible solution has been found. Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
However, this comes with the drawbacks of reduced interpretability, and similar to traditional spatial planning approaches such as Marxan, it relies on heuristics that can not guarantee that a best possible solution can been found [@Hanson2019OptimalityPrioritizing].

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning. 
Most software are based on linear or constrained programming approaches and require a solver to create any outputs.
<!-- R1: Be careful with such statements, it is extremely complex to objectively compare two solvers, and almost impossible to state the general superiority of one state-of-the-art solver against another one. Indeed, some solvers may be very efficient for some kinds of problems while very inefficient for others. Also, modelling and solver parametrization can make a lot of differences in solving efficiency and has often a lot of impact on performances. They are solver competitions, however, they are based on benchmarks, therefore classic and well-known problems that do not always reflect common uses of solvers. A solver fine-tuned to be good on such benchmarks will not necessarily perform well on any problem. There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
<!-- MJ: Agree to some extent, although again there are many comparisons that have shown that commercial solvers will usually outcompete open-source solvers.--->
Unfortunately, there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing] and the performance of any individual solvers can depend on the specific problem, data and computational resources.
Nevertheless, existing comparisons often show that proprietary solvers can be used to obtain solutions to complex problems in reasonable time periods [@Hanson2024SystematicConservation], however, due to software licenses and related cost certain solvers might not always be openly accessible.
